Causal Bandits: Learning Good Interventions via Causal Inference

### 2016-02-05 causal-bandit/icml-2015/general.tex
reworks intro to general graphs

We now consider the more general problem where the graph structure is known, but arbitrary. For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation). However if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorization formula \cite{Pearl2000}. 
\eq{
\P{X_1...X_N|do(X_i=j)} = 
\prod_{k \neq i}\P{X_k|\parents{X_k}}\delta(X_i - j)  
} 
Where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the dirac delta function. For example, given the causual graph in Figure \ref{fig:causalStructure_confounded}, by applying the truncated factorization and marginalizing, we obtain $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1}$.  Note that the quality of any estimator we build for $P{Y|do(X_2= j)}$ is not just a function of the number of times we observe $X_2 = j$.

@@

It is no longer optimal to divide our samples between observing to learn about multiple actions and explicit intervention to learn single ones. We need to incorporate the information we can learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ 
for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$.

@@

Let $P_a$ be a shorthand for $\P{.|a}$ and let $\eta$ be a distribution on available interventions $\calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.
Define $Q = \sum_{a \in \calA} \eta_a \Ps_a$ to be the mixture distribution over the interventions with respect to $\eta$.
Our algorithm will choose $T$ samples from $Q$ and use them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via
a truncated importance weighted estimator.

>>>>

We now consider the more general problem where the graph structure is known, but arbitrary. For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation). However if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorization formula \cite{Pearl2000}. 
\eq{
\P{X_1...X_N|do(X_i=j)} = 
\prod_{k \neq i}\P{X_k|\parents{X_k}}\delta(X_i - j)  
} 
Where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the dirac delta function. For example, given the causual graph in Figure \ref{fig:causalStructure_confounded}, by applying the truncated factorization and marginalizing, we obtain $\P{Y|do(X_2= j)} = \sum_{X_1}\P{Y|X_1,X_2=j}\P{X_1}$.

@@

We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions $a$ for which the observational estimates were poor. However, the variance of the observational estimator for $a = do(X_i = j)$ can be high even if $\P{X_i = j}$ is large. In our example, suppose $\P{X_1=\frac{1}{2}}$ and $X_2 = X_1$ deterministically. $\P{X_2 = 1} = \frac{1}{2}$, however we will never observe $(X_2=1,X_1 = 0)$ so we cannot get a good estimate for $\P{Y|do(X_2=1)}$.

Additionally, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable from rounds in which we act on a different variable. Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ 
for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$, (but not from $do()$).

@@

To overcome these difficulties we simplify the problem by assuming the conditional interventional distributions $\P{\boldsymbol{X}|a}$ are known for all variables except $Y$. Let $P_a$ be a shorthand for $\P{.|a}$ and let $\eta$ be a distribution on available interventions $\calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.
Define $Q = \sum_{a \in \calA} \eta_a \Ps_a$ to be the mixture distribution over the interventions with respect to $\eta$.
Our algorithm will choose $T$ samples from $Q$ and use them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via
a truncated importance weighted estimator.

### 2016-02-05 causal-bandit/icml-2015/introduction.tex
reworks intro to general graphs

Two pieces of recent work also consider using causal models in bandit problems.
\citet{Bareinboim2015} use causal inference to improve the chance of choosing an optimal arm. Their work differs from ours in two key ways. Firstly, their focus is on correcting the effects of unobserved confounding variables whereas we assume all variables are observable. Secondly, they do not assume their learning algorithm has access to a causal graph and instead use a form of contextual randomization to improve reward estimates in a Thompson sampling-style algorithm. \citet{Ortega2014thompson} present an analysis and extension of Thompson sampling assuming actions are causal interventions. Their focus is on causal induction (\ie, learning an unknown causal model) instead of exploiting a known causal model. Neither work provides a regret analysis. 
Combining their handling of unobserved variables and causal induction with our analysis is left as future work.
% Key to Elias' paper is: observing the action an agent would take if it were allowed to make its natural choice can provide some information about hidden confounders that influence both the reward and the choice of action. Therefore, incorporating an agents natural choice as context may outperform a standard bandit that does not use that context. (Note: even in the presence of hidden confounders, including the agents natural choice as context only may improve the results. It is easy to come up with a counter example in which it does not).

>>>>

Two pieces of recent work also consider applying ideas from causal inference to bandit problems.
\citet{Bareinboim2015} demonstrate that in the presence of confounding variables the value that a variable would have taken had it not been intervened on can provide important contextual information. \todof{write one sentence to explain how their work is different to ours}

 \citet{Ortega2014thompson} present an analysis and extension of Thompson sampling assuming actions are causal interventions. Their focus is on causal induction (\ie, learning an unknown causal model) instead of exploiting a known causal model. Combining their handling of  causal induction with our analysis is left as future work.
% Key to Elias' paper is: observing the action an agent would take if it were allowed to make its natural choice can provide some information about hidden confounders that influence both the reward and the choice of action. Therefore, incorporating an agents natural choice as context may outperform a standard bandit that does not use that context. (Note: even in the presence of hidden confounders, including the agents natural choice as context only may improve the results. It is easy to come up with a counter example in which it does not).

### 2016-02-05 causal-bandit/icml-2015/simple-regret.tex
reworks intro to general graphs

The algorithm operates as follows. For the first $T/2$ rounds it chooses $do()$ to collect observational data.
The fact that only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one means that $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$. Thus we can create good estimators of the returns of those actions $do(X_i = j)$ for which $\P{X_i = j}$ is large. 
These are the interventions that are easily estimated ``passively'' (\ie, without playing them explicitly). 
Those actions where $\P{X_i = j}$ is small may not be observed (often) in the first $T/2$ rounds, so
the estimates of the returns of these actions from the first $T/2$ samples could be quite poor. For the remaining $T/2$ rounds
the algorithm makes interventions on the actions $a = do(X_i = j)$ for which $\P{X_i = j}$ is small. The precise allocation is
carefully chosen to minimise the worst-case simple regret. 
The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables have an infrequently occurring value (\ie, small $q_i$ or $1-q_i$). 
Let $I_1 = \emptyset$ and for $\tau \in [0,1)$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} \leq \tau}$.
Define
\eqn{
\label{eq:m-simple}
m(\vec{q}) = \max \set{ d : |I_{\frac{1}{d}}| \leq d}\,.
% m(\vec{q}) = \max \set{i : \sum_{k=1}^N \ind{\min\set{q_k, \,1 - q_k} \leq 1/i} \leq i}\,.
}
This captures a trade off between identifying the low probability variables and not having too many of them.
When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 1$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$, the number of variables.

>>>>

The algorithm operates as follows. For the first $T/2$ rounds it chooses $do()$ to collect observational data.
As the only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one, $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$. We use this to get good estimates for the actions $do(X_i=j)$ for which $\P{X_i = j}$ is large, without playing them explicitly.

Those actions where $\P{X_i = j}$ is small may not be observed (often) in the first $T/2$ rounds, so the estimates of the returns of these actions from the first $T/2$ samples could be quite poor. For the remaining $T/2$ rounds the algorithm makes interventions on the actions $a = do(X_i = j)$ for which $\P{X_i = j}$ is small. The precise allocation is carefully chosen to minimise the worst-case simple regret. The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables have an infrequently occurring value (\ie, small $q_i$ or $1-q_i$). 
Let $I_1 = \emptyset$ and for $\tau \in [0,1)$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} \leq \tau}$.
Define
\eqn{
\label{eq:m-simple}
m(\vec{q}) = \max \set{ d : |I_{\frac{1}{d}}| \leq d}\,.
% m(\vec{q}) = \max \set{i : \sum_{k=1}^N \ind{\min\set{q_k, \,1 - q_k} \leq 1/i} \leq i}\,.
}
This captures a trade off between identifying the low probability variables and not having too many of them.
When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 1$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$, the number of variables.
